Comparing summed probability distributions of shoreline and radiocarbon dates from the Mesolithic Skagerrak coast of Norway

### 2023-08-22 analysis/paper/paper.Rmd
Minor changes before sending off to David

The potentially immense value of insights into population dynamics, combined with the ubiquity of radiocarbon dates and the relative ease with which these can now be processed within what has been termed a dates-as-data methodology [@rick1987] has undoubtedly contributed to the popularity of the approach. Several limitations and forms of criticism have, however, been directed at these procedures. Some of the objections are of a methodological nature, while others pertain to the underlying logic and the degree to which there is likely to be a constant and direct connection between the frequency of ^14^C-dates and population dynamics [e.g. @attenbrow2015; @carleton2021; @torfing2015; @ward2021]. What appears to be agreed upon by practitioners and critics alike is that radiocarbon dates are best analysed in this manner when compared and contrasted to other proxies for past population dynamics, and other variables that might impact these [@attenbrow2015; @crombe2014; @bevan2017; @french2021; @palmisano2017; @palmisano2021; @williams2012]. In this paper, we aim to report on and begin to unpack the relationship between two measures that have been linked to relative population size in the context of the Mesolithic Skagerrak coast in south-eastern Norway, namely the summed probability distribution of calibrated radiocarbon dates (RSPD) and the summed probability distribution of shoreline dated sites (SSPD). We discuss how the data compare with each other as well as how they can potentially be drawn on to understand cultural historical developments in the region through the Mesolithic, c. 9500--4000 BCE.

@@

The present study is decidedly exploratory. While some thoughts concerning the relationship between the variables underlie the analysis, these could not be instantiated as concrete hypotheses. Rather, we explore the data by using a set of standard models to begin to unpack patterns and grasp the relationship between the variables. The dates-as-data approach is dependent on there being a link between how prehistoric people generated material that become ^14^C samples and population dynamics [cf. @freeman2018]. The sum of shoreline dated sites, on the other hand, is determined by site frequency, and its use as a population proxy is therefore dependent on there being a connection between site count and population density. If a comparison of these proxies do not, or only partially correspond, it is thus an open question what factors have impacted either distribution to cause this discrepancy, and which measure, if any, reflects true population dynamics. The issue will therefore initially demand an open and exploratory approach where a multitude of explanatory and confounding effects can be drawn on to suggest explanations of any observed pattern.

@@

To what degree the radiocarbon record is determined by past population numbers might vary geographically and chronologically, based on variation in investigatory and taphonomic factors [@surovell2009b; @bluhm2019], as well as cultural processes within prehistoric populations. One example of the latter is the difference that might exist between farmer and forager populations, where @freeman2018 have suggested that an increased per capita energy consumption introduced with farming means that ^14^C-dates should not be weighted equally when making relative population estimates across such populations. Similarly, while site counts have also been invoked for the analysis of past population dynamics, these are likely to be impacted by factors such as land-use and mobility patterns, and settlement nucleation and dispersion, as well as variation in archaeological investigations and what has been taken to constitute a site [@drennan2015, 16--20; @palmisano2017]. However, these could be considered theoretical issues that have implications for how fluctuations in these proxies should be interpreted, and possibly weighted differently according some criteria. Before any such fluctuations are given any substantive interpretation, however, there is a host of methodological issues that have to be considered [see @carleton2021; @crema2022 for reviews].

@@

The most commonly applied of these is a null-hypothesis significance testing approach by means of Monte Carlo simulation, first introduced by @shennan2013 and later expanded upon by @timpson2014 and @crema2021. This works by comparing the observed RSPD with a series of simulated RSPDs, generated from a null model. These null models are typically a uniform, exponential or logistic distribution. These are chosen *a priori* with reference to common long-term population dynamics and are typically parametrised by fitting the models by means of least-squares or, given recent methodological developments, maximum likelihood estimation (MLE). The result from these simulations are then used to create a 95% critical envelope representing the null model. The proportion of the observed RSPD that falls outside the simulated envelope (i.e. positive or negative divergence of the observed SPD from the fitted null model), is then used to estimate a global p-value indicating whether the null model can be rejected. In the case that it can, the portions of the observed RSPD that falls outside this envelope can subsequently be interpreted as representing potentially meaningful demographic events, relative to the null model. However, care has to be taken in how these are interpreted. First, this follows from the fact that 5% of the deviations from the critical envelope can be expected to be random, and there is no way to know which deviations this pertains to. Secondly, a for example exponential null model fit to the data is only one of an infinite set of exponential models with different growth rates that could be used. While a model fit by means of MLE will likely have a reasonable growth rate, and by extension exclude many other exponential fits as likely to explain the data, this can be difficult to determine. Finally, the p-value only indicates whether or not the null model as a whole can be rejected as an explanation of the data, and does not provide statistical justification for inferring whether local deviations themselves are meaningful demographic signals [see review by @crema2022; @timpson2021].

The procedure outlined above represent the standard approach in efforts to model population dynamics based on RSPDs. This can be helpful for getting an impression of the overall developments of population proxies, and is an especially useful first step in the treatment of shoreline dated sites, as few assumptions concerning this overall development could be made. However, standard models frequently used to evaluate RSPDs are consistently being rejected in the literature and, it is worth underscoring, deviations from these rejected null models do not in themselves give statistical justification for the interpretation of any local deviations, nor a subsequent direct analysis of the SPDs themselves.

@@

Sites surveyed by means of test-pitting between Horten and Arendal were initially retrieved from the national heritage database Askeladden [@directorate2018], totalling at `r nrow(surveyed)` records. The records were then manually reviewed and given a quality score based on the criteria in Table \@ref(tab:tab1), indicating the degree to which the spatial location and extent of the sites is believed to be represented by the geometries available in the database [see also @roalkvam2020]. All sites with a quality score of 4 or worse were excluded from further analysis. Any sites situated at elevations that result in a shoreline date earlier than 9465 BCE were then excluded. This marks the latest start-date among the employed displacement curves, and no sites have yet been verified to have an older date than around 9300 BCE in the region [@glorstad2016]. Data on excavated sites was originally compiled for @roalkvam2023 and has been compared with site data as listed in @damlien2021 and @nielsen2019. Only excavated sites with available spatial data in Askeladden or local databases at the Museum of Cultural History of the University of Oslo were included in the analysis. The 102 excavated sites in the dataset without relevant ^14^C-dates that were originally shoreline dated in the excavation reports were included in the SSPD along with the retained surveyed sites. This gave a total of 921 shoreline dated sites in the final SSPD.

@@

All analyses done in this study were performed using the R programming language [@rcoreteam]. Underlying data and programming code used for the paper is available in a version-controlled online repository at https://doi.org/10.17605/osf.io/fqwt5. This is structured as a research compendium following @marwick2018b, to allow for reproducibility of the results [see also @Marwick2017]. Analysis of the ^14^C-dates were done using the R packages *ADMUR* [@timpson2021] and *rcarbon* [@crema2021]. The R package *shoredate* was used for performing and handling the shoreline dating of sites [@roalkvam2023b], with re-purposed code from both *ADMUR* and *rcarbon* for Monte Carlo simulation and model comparison.

@@

The analysis in @roalkvam2023 found that site phases ^14^C-dated to before 2500 BCE have a likely elevation above sea-level that can be reasonably approximated by a gamma distribution with shape =  0.286 and scale = 20.833. This is incorporated in the procedure for shoreline dating used for this study (Figure \@ref(fig:shoredate)D). When performing this dating procedure, the gamma distribution is sequentially stepped through and transferred to the calendar scale by uniformly distributing the probability across the years in the range between the lower and upper limit of the interpolated displacement curve. This procedure thus gives the shoreline date of a site when accounting for the likely elevation of the site above sea-level when it was in use [see @roalkvam2023 for details]. Given that the shoreline displacement curves have no inversions and should therefore before commutative [cf. @weninger2015, 545], each shoreline date is normalised to sum to unity. To reduce the computational cost of the simulation procedures to follow below, the gamma distribution is here stepped through at increments of 0.1 m and the calendar scale is kept at a resolution of 5 years.

@@

Although there are generally few ^14^C-dates from the earliest part of the Mesolithic in Norway [e.g. @åstveit2018; @kleppe2018; @breivik2018a; @solheim2018], it is established that Stone Age sites in southern Norway date to the period from before the earliest ^14^C-dates in the study area for this paper, and at least from as far back as around 9300 BCE [e.g. @damlien2018b; @glorstad2016]. While the SSPD achieved here starts from c. 9500 BCE, the RSPD shows no evidence of human activity until c. 8500 BCE. What has caused the lack of ^14^C-dates in south-eastern Norway from this earliest period has been given many possible explanations through the years. These range from taphonomic loss to diverse environmental and/or cultural factors, such as the use of shrubs or rapidly deteriorating seal blubber as fuel for burning, or a preference for consuming raw, dried or fermented food [see @bjerck2017; @damm2022]. A possible lack of reliable access to firewood has also been seen as possible contributor to this pattern, although evidence for the early presence of forested landscapes, at least in the southern parts of the country, has now rendered this a less favoured hypothesis [@solheim2018; cf. @damm2022; @sørensen2014]. While taphonomic and ecological factors are likely part of the picture, we view it as probable that the rapid growth at the start of RSPD and the magnitude of the SSPD in the period preceding this suggest that cultural and demographic factors are central drivers behind this pattern.

@@

Following this earliest phase of rapid increase in shoreline dated sites, there is a drop from this initial peak, starting from around 8800 BCE, concluding around 8500 BCE and remaining within the simulation envelope until c. 7900 BCE. It has previously been suggested that a change in mobility patterns occurs some time around 8500--8000 BCE, involving a reduction in residential mobility [e.g. @damlien2018b; @roalkvam2022]. Furthermore, both genetic data and technological analyses of lithic inventories indicate that the initial phase of human colonisation from the south is followed by an influx, and possible mixing of people from the east around the same time [@günther2018; @manninen2021]. If this narrative is coupled with the data observed here, these migratory events could appear to result in a relatively sudden drop in site-count as indicated in the SSPD, which in turn could be related to a change in mobility patterns. The end of this drop also roughly corresponds to the first ^14^C-dates in the RSPD around 8500 BCE, leading to the logistic growth that plateaus in the first centuries after 8000 BCE. The plateau of the logistic growth in the RSPD also has a corresponding positive deviation in the SSPD from the exponential model -- an increase that was also noted by @berg-hansen2022 -- which could indicate that this demographic signal is reflected in the frequency of shoreline dated sites as well.

@@

A recent study by @manninen2023, which considers a larger inland area of northern Scandinavia than the above-referenced studies found a corresponding increase in inland activity starting from the earliest parts of the Mesolithic that lasted until c. 6000 BCE. However, after this the population is found to have been characterised by a cessation of growth, rather than the clear decrease identified by the regional studies in southern Norway. Consequently, properly relating the developments in the coastal region to inland developments and their relevance to each other will benefit from an integrated approach that also explicitly considers the spatial dimension without invoking predefined analytical regions [cf. @crema2017].

Assessing the suggested and possibly varying influence of demographic developments and mobility patterns on the two employed proxies is challenging given the divergence between them without invoking additional variables that can . Following from the stationarity of the logistic model of the RSPD after c. 7500 BCE, it is possible that fluctuations in the SSPD reflect variation in mobility patterns that are unrelated to demographic developments. There does appear to be a negative deviation from the logistic null model in the RSPD corresponding to the one in the SSPD just before 6000 BCE. However, the logistic model could not be rejected for the RSPD, and so it is inappropriate to interpret any deviations from the Monte Carlo envelope and place any analytical weight on the deviations in the RSPD, at least given the present data. While it is possible that the fairly limited sample of ^14^C-dates has led to an underestimation of this deviation, there is, additionally, no indications in the RSPD of an increased demographic signal in the period just before 5000 BCE [see also @solheim2018]. A related point that is also worth making here is that the overall demographic trends that the modelling efforts can hope to identify are likely to have been subject to smaller scale demographic fluctuations that simply cannot be ascertained given the available data and its resolution [cf. @tallavaara2021], and which impacts the comparison of the two proxies following both from the different resolutions of the dating methods and the sample sizes underlying the SPDs.

Given the presence of people and corresponding signal in the SSPD before the start of the RSPD, and the suggested relevance of mobility patterns for the lacking signal in the RSPD in this earliest period, neither proxy is presumably devoid of influence from population density or mobility patterns. This is also given support by the correspondence between the logistic growth in the RSPD and the positive deviation in the SSPD in the first centuries of the 8th millennium BCE. However, properly assessing this relationship and the degree to which it has held constant throughout the period will depend on independent data that track the developments of the other factors that have been suggested to be of relevance to the behaviour of the two proxies.

@@

This paper has presented a new methodology for assembling and assessing the sum of a large number of shoreline dates. This was done by employing the probabilistic method for shoreline dating presented in @roalkvam2023, and by drawing on analogous procedures in the treatment of the temporal frequency distribution of radiocarbon dates [@crema2022; @timpson2014; @crema2021]. The result of employing this methodology was then compared to radiocarbon dates from within the same area, to begin to assess the relationship between the two data types and their potential association with overall developments in population density and mobility patterns.

The main results was that the RSPD is consistent with the logistic model fit to the data. This is characterised by a rapid logistic growth from the start of the RSPD, which plateaus and remains stationary from around 7500 BCE. This is contrasted to the shoreline dates, the development of which were better described by a function of exponential decay. While this indicates that some process of decline in the number of shoreline dated sites has occurred over time, the model was rejected as an adequate description of the SSPD. Some of the deviations from the model were then narratively explored to suggest some possible reasons for these deviations. However, given the lack of a model that adequately describe the data, and the informal treatment of these deviations, these should be viewed as suggestions that represent potential lines for further interrogation.

>>>>

The potentially immense value of insights into past population dynamics, combined with the ubiquity of radiocarbon dates and the relative ease with which these can now be processed within what has been termed a dates-as-data methodology [@rick1987] has undoubtedly contributed to the popularity of the approach. Several limitations and forms of criticism have, however, been directed at these procedures. Some of the objections are of a methodological nature, while others pertain to the underlying logic and the degree to which there is likely to be a constant and direct connection between the frequency of ^14^C-dates and population dynamics [e.g. @attenbrow2015; @carleton2021; @torfing2015; @ward2021]. What appears to be agreed upon by practitioners and critics alike is that radiocarbon dates are best analysed in this manner when compared and contrasted to other proxies for past population dynamics, and other variables that might impact these [@attenbrow2015; @crombe2014; @bevan2017; @french2021; @palmisano2017; @palmisano2021; @williams2012; @rick1987]. In this paper, we aim to report on and begin to unpack the relationship between two measures that have been linked to relative population size in the context of the Mesolithic Skagerrak coast in south-eastern Norway, namely the summed probability distribution of calibrated radiocarbon dates (RSPD) and the summed probability distribution of shoreline dated sites (SSPD). We discuss how the data compare with each other as well as how they can potentially be drawn on to understand cultural historical developments in the region through the Mesolithic, c. 9500--4000 BCE.

@@

The present study is decidedly exploratory. While some thoughts concerning the relationship between the variables underlie the analysis, these could not be instantiated as concrete hypotheses. Rather, we explore the data by using a set of standard models to begin to unpack patterns and grasp the relationship between the variables. The dates-as-data approach is dependent on there being a link between how prehistoric people generated material that become ^14^C samples and population dynamics. The sum of shoreline dated sites, on the other hand, is determined by site frequency, and its use as a population proxy is therefore dependent on there being a connection between site count and population density. If a comparison of these proxies do not, or only partially correspond, it is thus an open question what factors have impacted either distribution to cause this discrepancy, and which measure, if any, reflects true population dynamics. The issue will therefore initially demand an open and exploratory approach where a multitude of explanatory and confounding effects can be drawn on to suggest explanations of any observed pattern.

@@

To what degree the radiocarbon record is determined by past population numbers might vary geographically and chronologically, based on variation in investigatory and taphonomic factors [@surovell2009b; @bluhm2019], as well as cultural processes within prehistoric populations. One example of the latter is the difference that might exist between farmer and forager populations, where @freeman2018 have suggested that an increased per capita energy consumption introduced with farming means that ^14^C-dates should not be weighted equally when making relative population estimates across such populations. Similarly, while site counts have also been invoked for the analysis of past population dynamics, these are likely to be impacted by factors such as land-use and mobility patterns, and settlement nucleation and dispersion, as well as variation in archaeological investigations and what has been taken to constitute a site [@drennan2015, 16--20; @palmisano2017]. However, these could be considered theoretical issues that have implications for how fluctuations in these proxies should be interpreted, and possibly weighted differently according some criteria. Before any such fluctuations are given any substantive interpretation, however, there is a host of methodological issues that have to be considered [see @crema2022 for a recent review].

@@

The most commonly applied of these is a null-hypothesis significance testing approach by means of Monte Carlo simulation, first introduced by @shennan2013 and later expanded upon by @timpson2014 and @crema2021. This works by comparing the observed RSPD with a series of simulated RSPDs, generated from a null model. These null models are typically a uniform, exponential or logistic distribution. These are chosen *a priori* with reference to common long-term population dynamics and are typically parametrised by fitting the models by means of least-squares or, given recent methodological developments [@timpson2021], maximum likelihood estimation (MLE). The result from these simulations are then used to create a 95% critical envelope representing the null model. The proportion of the observed RSPD that falls outside the simulated envelope (i.e. positive or negative divergence of the observed SPD from the fitted null model), is then used to estimate a global p-value indicating whether the null model can be rejected. In the case that it can, the portions of the observed RSPD that falls outside this envelope can subsequently be interpreted as representing potentially meaningful demographic events, relative to the null model. However, care has to be taken in how these are interpreted. First, this follows from the fact that 5% of the deviations from the critical envelope can be expected to be random, and there is no way to know which deviations this pertains to. Secondly, a for example exponential null model fit to the data is only one of an infinite set of exponential models with different growth rates that could be used. While a model fit by means of MLE will likely have a reasonable growth rate, and by extension exclude many other exponential fits as likely to explain the data, this can be difficult to determine. Finally, the p-value only indicates whether or not the null model as a whole can be rejected as an explanation of the data, and does not provide statistical justification for inferring whether local deviations themselves are meaningful demographic signals [@crema2022; @timpson2021].

The procedure outlined above represent the standard approach in efforts to model population dynamics based on RSPDs. This can be helpful for getting an impression of the overall developments of population proxies, and is an especially useful first step in the treatment of shoreline dated sites, as few assumptions concerning this overall development could be made. However, it is worth underscoring that deviations from these rejected null models do not in themselves give statistical justification for the interpretation of any local deviations, nor a subsequent direct analysis of the SPDs themselves.

@@

Sites surveyed by means of test-pitting between Horten and Arendal were initially retrieved from the national heritage database Askeladden [@directorate2018], totalling at `r nrow(surveyed)` records. The records were then manually reviewed and given a quality score based on the criteria in Table \@ref(tab:tab1), indicating the degree to which the spatial location and extent of the sites is believed to be represented by the geometries available in the database [see also @roalkvam2020]. All sites with a quality score of 4 or worse were excluded from further analysis. Any sites situated at elevations that result in a shoreline date earlier than 9465 BCE were then excluded. This marks the latest start-date among the employed displacement curves, and no sites in the region have yet been conclusively verified to have an older date than around 9300 BCE [@glorstad2016]. Data on excavated sites was originally compiled for @roalkvam2023 and has been compared with site data as listed in @damlien2021 and @nielsen2019. Only excavated sites with available spatial data in Askeladden or local databases at the Museum of Cultural History of the University of Oslo were included in the analysis. The 102 excavated sites in the dataset without relevant ^14^C-dates that were originally shoreline dated in the excavation reports were included in the SSPD along with the retained surveyed sites. This gave a total of 921 shoreline dated sites in the final SSPD.

@@

All analyses done in this study were performed using the R programming language [@rcoreteam]. Underlying data and programming code used for the paper is available in a version-controlled online repository at https://doi.org/10.17605/osf.io/fqwt5. This is structured as a research compendium following @marwick2018b, to allow for reproducibility of the results [see also @Marwick2017]. Analysis of the ^14^C-dates were done using the R packages *ADMUR* [@timpson2021] and *rcarbon* [@crema2021]. The R package *shoredate* was used for performing and handling the shoreline dating of sites [@roalkvam2023b], and re-purposed code from both *ADMUR* and *rcarbon* for subsequent Monte Carlo simulation and model comparison.

@@

The analysis in @roalkvam2023 found that site phases ^14^C-dated to before 2500 BCE have a likely elevation above sea-level that can be reasonably approximated by a gamma distribution with shape =  0.286 and scale = 20.833. This is incorporated in the procedure for shoreline dating used for this study (Figure \@ref(fig:shoredate)D). When performing this dating procedure, the gamma distribution is sequentially stepped through and transferred to the calendar scale by uniformly distributing the probability across the years in the range between the lower and upper limit of the interpolated displacement curve. This procedure thus gives the shoreline date of a site when accounting for the likely elevation of the site above sea-level when it was in use [see @roalkvam2023 for details]. Given that the shoreline displacement curves have no inversions and should therefore before commutative [cf. @weninger2015, 545], each shoreline date is normalised to sum to unity before being summed. To reduce the computational cost of the simulation procedures to follow below, the gamma distribution is here stepped through at increments of 0.1 m and the calendar scale is kept at a resolution of 5 years.

@@

Although there are generally few ^14^C-dates from the earliest part of the Mesolithic in Norway [e.g. @åstveit2018; @kleppe2018; @breivik2018a; @solheim2018], it is established that Stone Age sites in southern Norway date to the period from before the earliest ^14^C-dates in the study area for this paper, and at least from as far back as around 9300 BCE [e.g. @damlien2018b; @glorstad2016]. While the SSPD achieved here starts from c. 9500 BCE, the RSPD shows no evidence of human activity until c. 8500 BCE. What has caused the lack of ^14^C-dates from this earliest period has been given many possible explanations through the years. These range from taphonomic loss to diverse environmental and/or cultural factors, such as the use of shrubs or rapidly deteriorating seal blubber as fuel for burning, or a preference for consuming raw, dried or fermented food [see @bjerck2017; @damm2022]. A possible lack of reliable access to firewood has also been seen as possible contributor to this pattern, although evidence for the early presence of forested landscapes, at least in the southern parts of the country, has now rendered this a less favoured hypothesis [@solheim2018; cf. @damm2022; @sørensen2014]. While taphonomic and ecological factors are likely part of the picture, we view it as probable that the rapid growth at the start of RSPD and the magnitude of the SSPD in the period preceding this suggest that cultural and demographic factors are central drivers behind this pattern.

@@

Following this earliest phase of rapid increase in shoreline dated sites, there is a drop from this initial peak, starting from around 8800 BCE, concluding around 8500 BCE and remaining within the simulation envelope until c. 7900 BCE. It has previously been suggested that a change in mobility patterns occurs some time around 8500--8000 BCE, involving a reduction in residential mobility [e.g. @damlien2018b; @roalkvam2022]. Furthermore, both genetic data and technological analyses of lithic inventories indicate that the initial phase of human colonisation from the south is followed by an influx, and possible mixing of people from the east around the same time [@günther2018; @manninen2021]. If this narrative is coupled with the data observed here, these migratory events could appear to result in a relatively sudden drop in site-count as indicated in the SSPD, which in turn could be related to a change in mobility patterns. The end of this drop also roughly corresponds to the first ^14^C-dates in the RSPD around 8500 BCE, leading to the logistic growth that plateaus in the first centuries after 8000 BCE. The logistic growth in the RSPD reaches its plateau in the same time-interval as a positive deviation in the SSPD from the exponential model -- an increase that was also noted by @berg-hansen2022 -- which could indicate that a demographic signal is reflected in the frequency of shoreline dated sites as well.

@@

A recent study by @manninen2023, which considers a larger inland area of northern Scandinavia than the above-referenced studies, found a corresponding increase in inland activity starting from the earliest parts of the Mesolithic that lasted until c. 6000 BCE. However, after this the population dynamics indicated by the radiocarbon record is found to have been characterised by a cessation of growth, rather than the clear decrease suggested by the regional studies in southern Norway. Consequently, properly relating the developments in the coastal region to inland developments and their relevance to each other will benefit from an integrated approach that also explicitly considers the spatial dimension without invoking predefined analytical regions [cf. @crema2017].

Following from the divergence between the proxies, assessing the suggested and possibly varying influence of demographic developments and mobility patterns on the measures will remain speculative without drawing on additional variables that can control for such confounding effects. Following from the stationarity of the logistic model of the RSPD after c. 7500 BCE, it is possible that fluctuations in the SSPD reflect variation in mobility patterns that are unrelated to demographic developments. There does appear to be a negative deviation from the logistic null model in the RSPD corresponding to the one in the SSPD just before 6000 BCE. However, the logistic model could not be rejected for the RSPD, and so it is inappropriate to interpret any deviations from the Monte Carlo envelope and place any analytical weight on the deviations in the RSPD, at least given the present data. While it is possible that the fairly limited sample of ^14^C-dates has led to an underestimation of this deviation, there is, additionally, no indications in the RSPD of an increased demographic signal in the period just before 5000 BCE [see also @solheim2018]. A related point that is also worth making here is that the overall demographic trends that the modelling efforts can hope to identify are likely to have been subject to smaller scale demographic fluctuations that simply cannot be ascertained given the available data and its resolution [cf. @tallavaara2021], and which impacts the comparison of the two proxies following both from the different resolutions of the dating methods and the sample sizes underlying the SPDs.

Given the presence of people and corresponding signal in the SSPD before the start of the RSPD, and the suggested relevance of mobility patterns for the lacking signal in the RSPD in this earliest period, neither proxy is presumably devoid of influence from population density or mobility patterns. This is also given support by the correspondence between the logistic growth in the RSPD and the positive deviation in the SSPD in the first centuries of the 8th millennium BCE. However, properly assessing this relationship and the degree to which it has held constant throughout the period will, as mentioned, depend on independent data that track the developments of the other factors that have been suggested to be of relevance to the behaviour of the two proxies.

@@

This paper has presented a new methodology for assembling and assessing the sum of a large number of shoreline dates. This was done by employing the probabilistic method for shoreline dating presented in @roalkvam2023, and by drawing on analogous procedures in the treatment of the temporal frequency distribution of radiocarbon dates [@crema2022; @timpson2014; @crema2021]. The result of employing this methodology was then compared to radiocarbon dates from within the same area, to begin to assess the relationship between the two data types.
<!-- and their potential association with overall developments in population density and mobility patterns. -->

The main results was that the RSPD is consistent with the logistic model fit to the data. This is characterised by a rapid logistic growth from the start of the RSPD, which plateaus and remains stationary from the first centuries after 8000 BCE. This is contrasted to the shoreline dates, the development of which were better described by a function of exponential decay. While this indicates that some process of decline in the number of shoreline dated sites has occurred over time, the model was rejected as an adequate description of the SSPD. Some of the deviations from the model were then narratively explored to suggest some possible reasons for these deviations. However, given the lack of a model that adequately describe the data, and the informal treatment of these deviations, these should be viewed as suggestions that represent potential lines for further interrogation.
